---
title: "The 2024 NFL season in 10 graphs"
output: html_notebook
---


```{r setup, include=TRUE, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
    tidyverse,
    nflverse,
    gt,
    gtExtras,
    gtUtils,
    ggfx,
    rvest,
    httr,
    jsonlite,
    patchwork,
    ggtext,
    ggrepel,
    ggimage,
    dlookr,
    showtext,
    sysfonts,
    janitor,
    grid,
    ggiraph,
    glue,
    here
)    

sysfonts::font_add_google("Lexend")
showtext::showtext_auto()
```

```{r nfl_analytics_theme, include=TRUE}
theme_nfl_analytics <- function(..., base_size = 12) {
  theme(
    text = element_text(family = "Lexend", size = base_size, color = "black", face='bold'),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16, vjust = .02, hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold", hjust = .5),
    strip.text = element_text(face = "bold"))
}
```

Potential ideas to explore:

* Benching Richardson
* Arthur Smith play action
* Dolphins: with and without Tua
* Regression of Stroud, Rodgers 
* Lions offensive efficiency
* Vikings defensive consistency 


#### Rookie QBs: performance and inconsistency

* Jayden Daniels spotlight

```{r}
pbp_2024 <- load_pbp(seasons = 2024)
ftn_2024 <- load_ftn_charting(seasons = 2024)

pbp_2024 <- pbp_2024 %>% 
  left_join(x=.,
            y=ftn_2024,
            by=c('game_id' = 'nflverse_game_id',
                 'play_id' = 'nflverse_play_id',
                 'season',
                 'week'))
```

```{r params, include=TRUE}
max_week_param <- pbp_2024 %>% 
  summarise(max(week)) %>% 
  pull()
```

```{r}
pbp_2024 %>% 
  select(contains('epa'))
```


```{r}
(rookie_qb_epa_weekly_2024 <- pbp_2024 %>% 
  filter(passer %in% c('J.Daniels', 'C.Williams', 'B.Nix', 'D.Maye')) %>% 
  group_by(game_id, week, posteam, passer, passer_id) %>% 
  summarise(across(
    c('qb_epa', 'pass_attempt', 'air_epa', 'yac_epa'),
    list(
        mean = ~round(mean(., na.rm = TRUE), 2),
        sum = ~round(sum(., na.rm = TRUE), 2)
      ),
      .names = "{.fn}_{.col}"
    ),
    .groups = 'drop') %>% 
  filter(sum_pass_attempt >= 10))
```

```{r}
rookie_qb_epa_weekly_2024 %>% 
  ggplot(.,
         aes(x=week,
             y=mean_qb_epa,
             color=posteam)) + 
  geom_line(linewidth=1.2) + 
  geom_point(data=. %>% filter(week == max_week_param),
             size=3) + 
  geom_text(data=. %>% filter(week == max_week_param),
            aes(x=week + 1,
                label=passer),
            family='Lexend') + 
  geom_hline(yintercept = 0, color='black', linetype='dashed') + 
  labs(title='EPA per dropback of 1st round rookies in 2024',
       subtitle = glue('Through Week {max_week_param}'),
       caption = 'Source: {nflverse} package<br>Minimum 10 attempts to qualify as a full game',
       x='Week',
       y='EPA/dropback') + 
  theme_nfl_analytics() +
  # theme(legend.position = 'top') + 
  scale_color_nfl(type='secondary')
  # theme(axis.title.y = element_text(angle=0))
```


```{r}
pbp_2024 %>% 
  # filter(passer %in% c('J.Daniels', 'C.Williams', 'B.Nix', 'D.Maye')) %>% 
  filter(!is.na(passer)) %>% 
  group_by(posteam, passer, passer_id) %>% 
  summarise(across(
    c('qb_epa', 'pass_attempt', 'air_epa', 'yac_epa'),
    list(
        mean = ~round(mean(., na.rm = TRUE), 5),
        sum = ~round(sum(., na.rm = TRUE), 5)
      ),
      .names = "{.fn}_{.col}"
    ),
    .groups = 'drop') %>% 
  filter(sum_pass_attempt >= 175) %>% 
  mutate(rank_qb_epa = rank(-mean_qb_epa)) %>% 
  filter(passer %in% c('J.Daniels', 'C.Williams', 'B.Nix', 'D.Maye')) %>%
  ggplot(.,
         aes(y=reorder(passer_id, mean_qb_epa),
             x=mean_qb_epa)) + 
  geom_col(aes(fill = if_else(mean_qb_epa >= 0,
                              '#81fd02', '#8f55e5')),
           width = 0.1) + 
  ggrepel::geom_text_repel(aes(x=mean_qb_epa,
                               label = glue('NFL rank: {rank_qb_epa}')), 
                           family='Lexend') + 
  labs(y='',
       x='EPA per dropback',
       title='EPA per dropback among rookie 1st-rounders in 2024', 
       subtitle=glue('Through Week {max_week_param}'), 
       caption='Source: {nflverse} package<br>Minimum 175 attempts') + 
  theme_nfl_analytics() + 
  theme(legend.position = 'none',
        axis.text.y = nflplotR::element_nfl_headshot(size = 3))
```

#### Kickoff eval

```{r}
kickoff_99_24 <- load_pbp(seasons = TRUE) %>% 
  filter(play_type == 'kickoff',
         season_type == 'REG')


kickoff_99_24 <- kickoff_99_24 %>%
  mutate(injury = if_else(str_detect(desc, 'injur'), 1, 0),
         injury_head = if_else(str_detect(desc, 'head injury'), 1, 0),
         injury_neck = if_else(str_detect(desc, 'neck injury'), 1, 0))

# what do I want the eval metrics to be?
# touchback
# injury
# return_yards
# kick_distance
# touchback
# kick_distance
# kickoff_inside_twenty
# kickoff_in_endzone
# kickoff_fair_catch
# touchdown

kickoff_vars <- c('touchback', 'return_yards', 'kick_distance', 'kickoff_inside_twenty',
                  'kickoff_in_endzone', 'kickoff_fair_catch', 'return_touchdown',
                  # it appears that injury coding changed and the desc field doesn't list injury type anymore 
                  'injury', 'injury_head', 'injury_neck')


kickoff_rates_99_24 <- kickoff_99_24 %>% 
  group_by(season) %>% 
  summarise(across(kickoff_vars,
                   list(
                     # mean = ~round(mean(., na.rm=TRUE), 5)
                     mean = ~mean(., na.rm=TRUE)
                   ),
                   .names = "{.fn}_{.col}"
                   ),
            .groups = 'drop') %>% 
  mutate(return_rate = 1 - mean_touchback)


plot_kickoff_var <- function(var){
  kickoff_rates_99_24 %>% 
  ggplot(.,
         aes(x=season,
             y= {{ var }})) + 
  geom_col() + 
  geom_mean_lines(data = . %>% filter(season != 2024),
                  aes(y0= {{ var }})) + 
  labs(x='Season',
       title="Evaluating the Impact of the NFL's Dynamic Kickoff Rule in 2024", 
       caption='Source: {nflverse}<br>Mean line excludes 2024 season') + 
  scale_y_continuous(labels = scales::percent_format(),
                     breaks = scales::pretty_breaks()) + 
  scale_x_continuous(breaks = scales::pretty_breaks()) + 
  theme_nfl_analytics()
}

plot_kickoff_var(mean_injury)
```



#### 2 shell and passing down

It doesn't seem like they put up the participation data during the season...

```{r}
participation_2023 <- load_participation(season = 2023)
```


#### Regression of Stroud, Rodgers 

```{r}
player_stats_99_24 <- load_player_stats(seasons = TRUE,
                                       stat_type = 'offense')

season_qbr_06_24 <- nflreadr::load_espn_qbr(seasons = TRUE,
                        summary_type = 'season')

avg_qbr <- season_qbr_06_24 %>% 
  filter(qualified == TRUE,
         season_type == 'Regular') %>% 
  summarise(median_qbr_total = median(qbr_total, na.rm=TRUE),
            .groups = 'drop') %>% 
  pull()



season_qbr_06_24 %>% 
  filter(name_short == 'A. Rodgers',
         season_type == 'Regular') %>% 
  ggplot(.,
         aes(x=season,
             y=qbr_total)) + 
  geom_line() + 
  geom_smooth(se=FALSE) + 
  geom_median_lines(data = season_qbr_06_24 %>% filter(qualified == TRUE,
                                                       season_type == 'Regular'),
                    aes(y0 = qbr_total)) + 
  scale_y_continuous(breaks = scales::pretty_breaks()) + 
  scale_x_continuous(breaks = scales::pretty_breaks()) + 
  theme_nfl_analytics()
```


```{r}
year <- 2024
pbp_test <- load_pbp(seasons = year)
pbp_test %>% 
  select(contains('epa'))

rodgers_pbp_08_24 <- load_pbp(seasons = c(2008:2022,2024)) %>% 
  filter(season_type == 'REG',
         !is.na(passer)) %>% 
  group_by(season, passer, posteam) %>% 
  summarise(plays = n(),
            # count_drops = sum(is_drop, na.rm = TRUE),
            interceptions = sum(interception, na.rm = TRUE),
            sum_qb_epa = sum(qb_epa, na.rm = TRUE),
            sum_air_epa = sum(air_epa, na.rm = TRUE),
            sum_yac_epa = sum(yac_epa, na.rm = TRUE),
            # sum_qb_epa_drops = sum(qb_epa[is_drop == TRUE], na.rm = TRUE),
            sum_qb_epa_int = sum(qb_epa[interception == 1], na.rm = TRUE),
            sum_qb_epa_sack = sum(qb_epa[sack == 1], na.rm = TRUE),
            # sum_qb_epa_incomplete = sum(qb_epa[incomplete_pass == 1], na.rm = TRUE),
            sum_qb_epa_dpi = sum(qb_epa[penalty_type == 'Defensive Pass Interference'], na.rm = TRUE),
            .groups = 'drop') %>% 
  filter(plays >= 25) %>% 
  mutate(
    across(
      starts_with("sum"),
      list(
        # mean = ~round(mean(., na.rm = TRUE), 2),
        # zscore = ~scale(.),
        mean_centered = ~. - mean(., na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    ))
```



```{r}
passing_epa_2024 <- pbp_2024 %>% 
  filter(!is.na(passer)) %>% 
  group_by(passer, posteam) %>% 
  summarise(plays = n(),
            count_drops = sum(is_drop, na.rm = TRUE),
            interceptions = sum(interception, na.rm = TRUE),
            sum_qb_epa = sum(qb_epa, na.rm = TRUE),
            sum_qb_epa_drops = sum(qb_epa[is_drop == TRUE], na.rm = TRUE),
            sum_qb_epa_int = sum(qb_epa[interception == 1], na.rm = TRUE),
            sum_qb_epa_sack = sum(qb_epa[sack == 1], na.rm = TRUE),
            sum_qb_epa_incomplete = sum(qb_epa[incomplete_pass == 1], na.rm = TRUE),
            sum_qb_epa_dpi = sum(qb_epa[penalty_type == 'Defensive Pass Interference'], na.rm = TRUE),
            .groups = 'drop') %>% 
  filter(plays >= 25) %>% 
  mutate(
    across(
      starts_with("sum"),
      list(
        # mean = ~round(mean(., na.rm = TRUE), 2),
        # zscore = ~scale(.),
        mean_centered = ~. - mean(., na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    )) %>% 
  arrange(-sum_qb_epa_int)
```


```{r}
# MVP years: 2014, 2020, 2021

rodgers_pbp_08_24 %>% 
  filter(passer == 'A.Rodgers') %>% 
  select(passer, season, posteam, plays, contains('mean_centered')) %>% 
  arrange(-season) %>% 
  gt() %>% 
  cols_label(passer = 'Name',
             season = 'Season',
             posteam = 'Team',
             plays = 'Plays',
             mean_centered_sum_qb_epa = 'EPA',
             mean_centered_sum_air_epa = 'Air',
             mean_centered_sum_yac_epa = 'YAC',
             mean_centered_sum_qb_epa_int = 'INTs',
             mean_centered_sum_qb_epa_sack = 'Sacks',
             mean_centered_sum_qb_epa_dpi = 'DPI') %>% 
  tab_style(
    style = list(
      cell_fill(color = "lightgray"),
      cell_text(weight = "bold")
    ),
    locations = cells_body(
      rows = season %in% c(2014, 2020, 2021)
    )
  ) %>% 
  tab_spanner(
    label = md("Mean Centered<br>Positive values indicate performance above league average in that season"),    
    columns = c(mean_centered_sum_qb_epa:mean_centered_sum_qb_epa_dpi)           # Columns to span
  ) %>% 
  gt_theme_athletic()
```



```{r qb_epa_mean_centered_gt, include=TRUE, warning=FALSE, results='asis'}
passing_epa_2024 %>% 
  select(passer, posteam, plays, contains('mean_centered'), -mean_centered_sum_qb_epa_incomplete) %>% 
  arrange(-mean_centered_sum_qb_epa) %>% 
  gt() %>% 
  cols_label(passer = 'Name',
             posteam = 'Team',
             plays = 'Plays',
             mean_centered_sum_qb_epa = 'EPA',
             mean_centered_sum_qb_epa_drops = 'Drops',
             mean_centered_sum_qb_epa_int = 'INTs',
             mean_centered_sum_qb_epa_sack = 'Sacks',
             mean_centered_sum_qb_epa_dpi = 'DPI') %>% 
  gt_theme_athletic()
```


