---
title: "NFL Analytics: a review of Expected Points Added"
output: html_notebook
---

```{r libraries, include=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nflverse)
library(gt)
library(gtExtras)
library(gtUtils)
library(ggtext)
library(ggrepel)
library(patchwork)
library(ggimage)
library(glue)
library(ggiraph)
```

```{r nfl_analytics_theme, include=TRUE}
nfl_analytics_theme <- function(..., base_size = 12) {
  
  theme(
    text = element_text(family = "Lexend",
                        size = base_size,
                        color = "black"),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16,
                                  vjust = .02,
                                  hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold", hjust = .5),
    strip.text = element_text(face = "bold"))
}
```

```{r teams, include=TRUE}
teams <- nflreadr::load_teams(current = TRUE)
```


This looks like a good article on [EPA](https://bestballstats.com/expected-points-added-a-full-explanation/#:~:text=Expected%20Points%20Added%20(EPA)%20is,evaluate%20team%20and%20player%20performance.) and [another](https://www.actionnetwork.com/education/what-is-epa-how-expected-points-added-advanced-metric-can-help-you-with-football-betting).


```{r play_by_play_2024, include=TRUE}
play_by_play_2024 <- load_pbp(seasons=most_recent_season())
```

# Offensive EPA 

```{r team_offensive_epa, include=TRUE}
(team_offensive_epa <- play_by_play_2024 %>% 
  group_by(posteam) %>% 
  filter(!is.na(posteam)) %>% 
  summarise(avg_offensive_epa = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  rename(team = posteam))
```

```{r team_defensive_epa, include=TRUE}
(team_defensive_epa <- play_by_play_2024 %>% 
  group_by(defteam) %>% 
  filter(!is.na(posteam)) %>% 
  summarise(avg_defensive_epa = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  rename(team = defteam))
```

```{r df_team_epa, include=TRUE}
(df_team_epa <- team_offensive_epa %>% 
  left_join(x=.,
            y=team_defensive_epa,
            by='team') %>% 
  left_join(x=.,
            y=teams,
            by=c('team' = 'team_abbr')))
```


What's the combined record of teams in each quadrant?

```{r}
df_team_epa %>% 
  ggplot(.,
         aes(x=avg_offensive_epa,
             y=avg_defensive_epa,
             alpha=if_else(team == 'PIT', 'PIT', 'Other'))) + 
  # geom_point(color=df_team_epa$team_color, size=3.5) + 
  geom_image(aes(image = team_logo_wikipedia), size=.1) +
  geom_hline(yintercept = 0,
             color = "black", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = 0,
             color = "black", linewidth = 0.8, linetype = "dashed") +
  scale_alpha_manual(values = c('PIT' = 1,
                                'Other' = .2)) + 
  scale_x_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  labs(x='Average Offensive EPA',
       y='Average Defensive EPA',
       caption='Source: nflreadr:: package') + 
  ggtitle('**Average Expected Points Added**: Offense vs. Defense',
          subtitle = '2024 season') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
```

```{r}
play_by_play_2024 %>% 
  filter(!is.na(posteam)) %>% 
  group_by(posteam) %>% 
  summarise(avg_epa_per_play = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  left_join(x=.,
            y=teams,
            by=c('posteam' = 'team_abbr')) %>% 
  ggplot(.,
         aes(x = avg_epa_per_play,
             y = reorder(posteam, avg_epa_per_play))) +
  geom_col(aes(fill = if_else(avg_epa_per_play >= 0,
                              "#2c7bb6", "#d7181c")),
           width = 0.2) +
  geom_vline(xintercept = 0, color = "black") +
  scale_fill_identity() +
  geom_image(aes(image = team_logo_wikipedia),
             asp = 16/9, size = .035) +
  labs(x='EPA per Play',
       y='',
       title = "**Offensive Expected Points Added per Play**",
       subtitle = "*2024 Regular Season*",
       caption = "*Inspired by An Introduction to NFL Analytics with R*") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  scale_y_discrete(expand = c(.05, 0)) +
  nfl_analytics_theme() +
  theme(axis.text.y = element_blank())
```


```{r}
play_by_play_2024 %>% 
  filter(!is.na(posteam)) %>% 
  group_by(defteam) %>% 
  summarise(avg_epa_per_play = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  left_join(x=.,
            y=teams,
            by=c('defteam' = 'team_abbr')) %>% 
  ggplot(.,
         aes(x = avg_epa_per_play,
             y = reorder(defteam, -avg_epa_per_play))) +
  geom_col(aes(fill = if_else(avg_epa_per_play >= 0,
                              "#d7181c", "#2c7bb6")),
           width = 0.2) +
  geom_vline(xintercept = 0, color = "black") +
  scale_fill_identity() +
  geom_image(aes(image = team_logo_wikipedia),
             asp = 16/9, size = .035) +
  labs(x='EPA per Play',
       y='',
       title = "**Defensive Expected Points Added per Play**",
       subtitle = "*2024 Regular Season*",
       caption = "*Inspired by An Introduction to NFL Analytics with R*") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  scale_y_discrete(expand = c(.05, 0)) +
  nfl_analytics_theme() +
  theme(axis.text.y = element_blank())
```


```{r epa_quadrant, include=TRUE}
(df_team_epa <- df_team_epa %>% 
mutate(epa_quadrant = if_else(avg_defensive_epa > 0 & avg_offensive_epa < 0, 'Quad1',
                      if_else(avg_defensive_epa > 0 & avg_offensive_epa > 0, 'Quad2',
                      if_else(avg_defensive_epa < 0 & avg_offensive_epa < 0, 'Quad3', 'Quad4')))))
```


```{r defensive_epa_runpass, include=TRUE}
(defensive_epa_runpass <- play_by_play_2024 %>% 
  filter(play_type %in% c('run', 'pass')) %>% 
  group_by(defteam, play_type) %>% 
  summarise(avg_epa = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  pivot_wider(names_from = play_type, values_from = avg_epa) %>% 
  left_join(x=.,
            y=teams,
            by=c('defteam' = 'team_abbr')) %>% 
    ggplot(.,
         aes(x=run,
             y=pass)) + 
  geom_image(aes(image = team_logo_wikipedia), size=.1) +
  geom_hline(yintercept = 0,
             color = "black", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = 0,
             color = "black", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  labs(x='Against the Run',
       y='Against the Pass',
       caption='Source: nflreadr:: package') + 
  ggtitle('**Average Defensive Expected Points Added**: Run vs. Pass',
          subtitle = '2024 season (through Week 7)') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none'))
```

```{r ftn_charting_2024, include=TRUE}
ftn_charting_2024 <- load_ftn_charting(seasons = most_recent_season())
```

```{r pbp_plus_ftn_2024, include=TRUE}
pbp_plus_ftn_2024 <- play_by_play_2024 %>% 
  left_join(x=.,
            y=ftn_charting_2024,
            by=c('game_id' = 'nflverse_game_id',
                 'play_id' = 'nflverse_play_id'))
```


```{r}
passing_epa <- pbp_plus_ftn_2024 %>% 
  filter(!is.na(passer)) %>% 
  group_by(passer, posteam) %>% 
  summarise(plays = n(),
            count_drops = sum(is_drop, na.rm = TRUE),
            interceptions = sum(interception, na.rm = TRUE),
            sum_qb_epa = sum(qb_epa, na.rm = TRUE),
            sum_qb_epa_drops = sum(qb_epa[is_drop == TRUE], na.rm = TRUE),
            sum_qb_epa_int = sum(qb_epa[interception == 1], na.rm = TRUE),
            sum_qb_epa_sack = sum(qb_epa[sack == 1], na.rm = TRUE),
            sum_qb_epa_incomplete = sum(qb_epa[incomplete_pass == 1], na.rm = TRUE),
            sum_qb_epa_dpi = sum(qb_epa[penalty_type == 'Defensive Pass Interference'], na.rm = TRUE),
            .groups = 'drop') %>% 
  filter(plays >= 25) %>% 
  mutate(
    across(
      starts_with("sum"),
      list(
        # mean = ~round(mean(., na.rm = TRUE), 2),
        # zscore = ~scale(.),
        mean_centered = ~. - mean(., na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    )) %>% 
  arrange(-sum_qb_epa_int)
```


```{r qb_epa_mean_centered_gt, include=TRUE, warning=FALSE, results='asis'}
passing_epa %>% 
  select(passer, posteam, plays, contains('mean_centered'), -mean_centered_sum_qb_epa_incomplete) %>% 
  arrange(-mean_centered_sum_qb_epa) %>% 
  gt() %>% 
  cols_label(passer = 'Name',
             posteam = 'Team',
             plays = 'Plays',
             mean_centered_sum_qb_epa = 'EPA',
             mean_centered_sum_qb_epa_drops = 'Drops',
             mean_centered_sum_qb_epa_int = 'INTs',
             mean_centered_sum_qb_epa_sack = 'Sacks',
             mean_centered_sum_qb_epa_dpi = 'DPI')
```


## Would the Broncos defensive EPA look dramatically different without Pat Surtain's 100 yard pick six?

```{r}
pbp_plus_ftn_2024 %>% 
  filter(game_id == '2024_05_LV_DEN', 
         play_id == 1192) %>% 
  select(contains('epa'))

```


```{r}
play_by_play_2024 %>% 
  filter(!(play_id == 1192 & game_id == '2024_05_LV_DEN')) %>% 
  filter(!is.na(posteam)) %>% 
  group_by(defteam) %>% 
  summarise(avg_epa_per_play = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  left_join(x=.,
            y=teams,
            by=c('defteam' = 'team_abbr')) %>% 
  ggplot(.,
         aes(x = avg_epa_per_play,
             y = reorder(defteam, -avg_epa_per_play))) +
  geom_col(aes(fill = if_else(avg_epa_per_play >= 0,
                              "#d7181c", "#2c7bb6")),
           width = 0.2) +
  geom_vline(xintercept = 0, color = "black") +
  scale_fill_identity() +
  geom_image(aes(image = team_logo_wikipedia),
             asp = 16/9, size = .035) +
  labs(x='EPA per Play',
       y='',
       title = "**Defensive Expected Points Added per Play**",
       subtitle = "*2024 Regular Season (through Week 7)<br>Excludes Pat Surtain's pick 6 to see its effect on value*",
       caption = "*Inspired by An Introduction to NFL Analytics with R*") + 
  scale_x_continuous(breaks = scales::pretty_breaks(n = 6)) +
  scale_y_discrete(expand = c(.05, 0)) +
  nfl_analytics_theme() +
  theme(axis.text.y = element_blank())
```


```{r defensive_epa_runpass_wo_surtain_pick6, include=TRUE}
(defensive_epa_runpass_wo_surtain_pick6 <- play_by_play_2024 %>% 
  filter(!(play_id == 1192 & game_id == '2024_05_LV_DEN')) %>% 
  filter(play_type %in% c('run', 'pass')) %>% 
  group_by(defteam, play_type) %>% 
  summarise(avg_epa = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  pivot_wider(names_from = play_type, values_from = avg_epa) %>% 
  left_join(x=.,
            y=teams,
            by=c('defteam' = 'team_abbr')) %>% 
    ggplot(.,
         aes(x=run,
             y=pass)) + 
  geom_image(aes(image = team_logo_wikipedia), size=.1) +
  geom_hline(yintercept = 0,
             color = "black", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = 0,
             color = "black", linewidth = 0.8, linetype = "dashed") +
  scale_x_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  labs(x='Against the Run',
       y='',
       caption='Source: nflreadr:: package') + 
  ggtitle('',
          subtitle = "<span style='color:#FB4F14'>**Excludes Pat Surtain's pick 6**</span>") + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none'))
```

```{r patchwork_runpass_wo_surtain_pick6, include=TRUE}
defensive_epa_runpass + defensive_epa_runpass_wo_surtain_pick6
```

This shows that, while that play did have an effect, the Broncos would still probably have the top defense against the pass through Week 7 (I say probably, because I didn't substitute that play's actual EPA with what it would have been, had that play resulted in a touchdown).

## Pat Mahomes' interception misfortune in 2024 -- what do our priors tell us about his likely performance from here? 

```{r}
passing_epa %>% 
  ggplot(.,
         aes(x=interceptions,
             y=sum_qb_epa_int,
             alpha=if_else(passer == 'P.Mahomes', 'P.Mahomes', 'Other'))) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = passer),
                           box.padding = .5) + 
  scale_x_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_alpha_manual(values = c('P.Mahomes' = 1,
                                'Other' = .2)) + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
```

```{r player_stats_17_24, include=TRUE}
player_stats_17_24 <- load_player_stats(seasons = 2017:2024,
                                        stat_type = 'offense')
```


```{r}
(int_rate_17_24 <- player_stats_17_24 %>% 
  filter(position == 'QB') %>% 
  group_by(recent_team, player_name, season) %>% 
  summarise(
    across(
      c(attempts, interceptions),
      list(
        sum = ~sum(., na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    ),
    .groups = 'drop') %>% 
  mutate(int_rate = sum_interceptions / sum_attempts) %>% 
  filter(sum_attempts >= 100))
```


```{r}
int_rate_17_24 %>% 
  filter(player_name == 'P.Mahomes') %>% 
  ggplot(.,
         aes(x=season,
             y=int_rate)) + 
  geom_col() + 
  geom_mean_lines(aes(y0=int_rate)) + 
  geom_text(aes(label = paste(round(int_rate*100, 1), '%', sep='')),
            nudge_y = -.002, color='white', family='Lexend', fontface='bold') + 
  scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = seq(2017, 2024, 1)) +
  labs(x='',
       y='INT rate') + 
  ggtitle('Pat Mahomes has **doubled** his career average interception rate',
          subtitle = 'INT rate = interceptions / attempts<br>Does not exclude throw aways') + 
  nfl_analytics_theme()
```

```{r int_rate_jitter, include=TRUE, fig.width=10, fig.height=6.2}
int_rate_jitter <- int_rate_17_24 %>% 
  filter(season >= 2018) %>% 
  group_by(season) %>% 
  mutate(rank_int_rate = rank(int_rate),
         is_mahomes = if_else(player_name == 'P.Mahomes', 'P.Mahomes', 'Other')) %>% 
  ggplot(.,
         aes(x=season,
             y=int_rate,
             color=is_mahomes,
             alpha=is_mahomes)) + 
  # geom_jitter(height = 0) + 
  geom_jitter_interactive(aes(data_id = player_name, tooltip = player_name),
                          height = 0) + 
  ggtitle(glue("<span style='color:#E31837'>**Pat Mahomes**</span> usually boasts some of the lowest INT rates<br>in the NFL -- not in 2024")) + 
  scale_color_manual(values = c('P.Mahomes' = '#E31837',
                                'Other' = 'black')) + 
  scale_alpha_manual(values = c('P.Mahomes' = 1,
                                'Other' = .2)) + 
  scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = seq(2017, 2024, 1)) +
  labs(x='',
       y='INT rate') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
```


```{r}
interactive_plot <- girafe(
  ggobj = int_rate_jitter
)
```

```{r int_rate_jitter_interactive, include=TRUE, fig.width=10, fig.height=6.2}
int_rate_jitter <- int_rate_17_24 %>% 
  filter(season >= 2018) %>% 
  group_by(season) %>% 
  mutate(rank_int_rate = rank(int_rate),
         is_mahomes = if_else(player_name == 'P.Mahomes', 'P.Mahomes', 'Other'),
         player_name_clean = stringr::str_replace_all(player_name, "'", "")) %>%  # Remove single quotes
  ggplot(aes(x = season,
             y = int_rate,
             color = is_mahomes,
             alpha = is_mahomes)) + 
  geom_jitter_interactive(aes(
    data_id = player_name_clean, 
    tooltip = glue("{player_name_clean} ({season}): {scales::percent(int_rate)}")
  ), height = 0) + 
  ggtitle(glue("Pat Mahomes usually boasts some of the lowest INT rates
in the NFL -- not in 2024")) + 
  scale_color_manual(values = c('P.Mahomes' = '#E31837', 'Other' = 'black')) + 
  scale_alpha_manual(values = c('P.Mahomes' = 1, 'Other' = 0.2)) + 
  scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = seq(2017, 2024, 1)) +
  labs(x = '', y = 'INT rate') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')

interactive_plot <- girafe(ggobj = int_rate_jitter)

interactive_plot
```



# QB EPA Tables 


```{r gt_epa_per_play, include=FALSE, warning=FALSE, results='asis'}
pbp_plus_ftn_2024 %>% 
  filter(!is.na(passer)) %>% 
  group_by(passer, posteam) %>% 
  summarise(sum_qb_epa = sum(qb_epa, na.rm = TRUE),
            sum_comp_yac_epa = sum(comp_yac_epa, na.rm = TRUE),
            plays = n(),
            .groups = 'drop') %>% 
  filter(plays >= 100) %>% 
  mutate(epa_per_play = round(sum_qb_epa / plays, 3),
         rank_epa_per_play = rank(-epa_per_play)) %>% 
  arrange(-epa_per_play) %>% 
  gt() %>% 
  cols_label(passer = 'Name',
             posteam = 'Team',
             sum_qb_epa = 'Total EPA',
             sum_comp_yac_epa = 'EPA from YAC (completions only)',
             plays = 'Plays',
             epa_per_play = 'EPA / Play',
             rank_epa_per_play = 'Rank') %>% 
  gt_color_pills(sum_qb_epa, digits = 3) %>% 
  gt_color_pills(sum_comp_yac_epa, digits = 3) %>% 
  gt_color_pills(plays, digits = 0) %>% 
  gt_color_pills(epa_per_play, digits = 3) %>% 
  tab_header(
    title = "Advanced Passing Stats: 2024 Season through Week 7",
    subtitle = "Sorted by Expected Points Added (EPA) per play. Minimum 100 plays."
  ) %>% 
  gt_theme_athletic() %>%
  cols_width(everything() ~ px(120)) %>% 
  gt_border_bars_top("#1c2632", bar_height = 5) %>%
  gt_border_bars_bottom("#1c2632", bar_height = 25, text = "Source: `nflreadr::` package. Inspiration: Georgios Karamanis' Tidy Tuesday work.", text_weight = "normal", text_size = 12)


qb_epa_mean_centered_gt <- 
  htmltools::withTags({
    passing_epa %>% 
  select(passer, posteam, plays, contains('mean_centered'), -mean_centered_sum_qb_epa_incomplete) %>% 
  arrange(-mean_centered_sum_qb_epa) %>% 
  gt() %>% 
  cols_label(passer = 'Name',
             posteam = 'Team',
             plays = 'Plays',
             mean_centered_sum_qb_epa = 'EPA',
             mean_centered_sum_qb_epa_drops = 'Drops',
             mean_centered_sum_qb_epa_int = 'INTs',
             mean_centered_sum_qb_epa_sack = 'Sacks',
             mean_centered_sum_qb_epa_dpi = 'DPI') %>% 
  gt_color_pills(mean_centered_sum_qb_epa, digits = 3) %>% 
  gt_color_pills(mean_centered_sum_qb_epa_drops, digits = 3) %>% 
  gt_color_pills(mean_centered_sum_qb_epa_int, digits = 3) %>% 
  gt_color_pills(mean_centered_sum_qb_epa_sack, digits = 3) %>% 
  gt_color_pills(mean_centered_sum_qb_epa_dpi, digits = 3) %>% 
  tab_header(
    title = "Advanced Passing Stats: 2024 Season through Week 7",
    subtitle = "All values relative to the average QB in 2024. Minimum 25 plays."
  ) %>% 
  gt_theme_athletic() %>%
  cols_width(everything() ~ px(120)) %>% 
  gt_border_bars_top("#1c2632", bar_height = 5) %>%
  gt_border_bars_bottom("#1c2632", bar_height = 25, text = "Source: `nflreadr::` package. Inspiration: Unexpected Points Substack.", text_weight = "normal", text_size = 12)
  })


```

```{r render_qb_epa_mean_centered_gt, include=TRUE}
qb_epa_mean_centered_gt
```