---
title: "Passing Breakdown with {nflverse}"
output: html_notebook
---

# Setup

```{r libraries, include=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nflverse)
library(gt)
library(gtExtras)
library(gtUtils)
library(ggtext)
library(ggrepel)
library(patchwork)
library(ggfx)
library(ggimage)
library(glue)
library(ggiraph)
```

```{r nfl_analytics_theme, include=TRUE}
nfl_analytics_theme <- function(..., base_size = 12) {
  
  theme(
    text = element_text(family = "Lexend",
                        size = base_size,
                        color = "black"),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16,
                                  vjust = .02,
                                  hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold", hjust = .5),
    strip.text = element_text(face = "bold"))
}
```

```{r teams, include=TRUE}
teams <- nflreadr::load_teams(current = TRUE)
```

```{r pbp_2022_24, include=TRUE}
pbp_2022_24 <- load_pbp(seasons=2022:2024) %>% 
  filter(season_type == 'REG')
```

FTN's charting (which is available in the {nflverse} package), contains additional play-level information that can be added to the `load_pbp()` data.

```{r ftn_charting_2022_24, include=TRUE}
ftn_charting_2022_24 <- load_ftn_charting(seasons = 2022:2024)
```

```{r pbp_plus_ftn_2022_24, include=TRUE}
pbp_plus_ftn_2022_24 <- pbp_2022_24 %>% 
  left_join(x=.,
            y=ftn_charting_2022_24,
            by=c('game_id' = 'nflverse_game_id',
                 'play_id' = 'nflverse_play_id',
                 'season'))
```


# Play action passing in Pittsburgh

```{r play_action, include=TRUE}
(play_action <- pbp_plus_ftn_2022_24 %>% 
  filter(play_type == 'pass',
         # sack == 0,
         qb_scramble == 0,
         !is.na(is_play_action)) %>% 
  group_by(season, posteam, is_play_action) %>% 
  summarise(plays = n(),
            sum_epa = sum(qb_epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  mutate(epa_per_play = sum_epa / plays) %>% 
  pivot_wider(names_from = is_play_action, values_from = plays:epa_per_play) %>% 
  mutate(percent_play_action = plays_TRUE / (plays_TRUE + plays_FALSE)) %>% 
  arrange(-percent_play_action) %>% 
  group_by(season) %>% 
  mutate(rank_percent_play_action = rank(-percent_play_action)) %>% 
  select(percent_play_action, everything()))

play_action %>% 
  filter(posteam == 'PIT')
```

```{r}
play_action_geom_col <- function(x){
  plot <- play_action %>% 
  filter(season == x) %>% 
  ggplot(.,
         aes(x=percent_play_action,
             y=reorder(posteam, percent_play_action),
             alpha=if_else(posteam == 'PIT', 'PIT', 'Other'))) + 
  geom_col() + 
  labs(x='Play Action Rate (%)',
       y='',
       title=x) + 
  scale_x_continuous(labels = scales::percent_format(),
                     limits = c(0, .4)) + 
  scale_fill_manual(values = c('PIT' = 1,
                               'Other' = .2)) + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
  
  return(plot)
}
```

```{r}
play_action_geom_col(2022) + play_action_geom_col(2023) + play_action_geom_col(2024)
```
Pittsburgh under Matt Canada finished in the bottom 4 or 3 teams with play action rate in 2022 and 2023, respectively. Midway through the 2024 season, Arthur Smith is using play action at over twice the rates seen under Matt Canada.

**Have these plays been successful?**

In a limited sample (two starts), **Russel Wilson is ranks 3rd among QBs** in EPA per play on play action passes.

```{r play_action_2024, include=TRUE}
(play_action_2024 <- pbp_plus_ftn_2022_24 %>% 
  filter(season == 2024,
         play_type == 'pass',
         qb_scramble == 0) %>% 
  group_by(posteam, passer_id, passer, is_play_action) %>% 
  summarise(plays = n(),
            sum_epa = sum(qb_epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  mutate(epa_per_play = sum_epa / plays) %>% 
  pivot_wider(names_from = is_play_action, values_from = plays:epa_per_play) %>% 
  mutate(percent_play_action = plays_TRUE / (plays_TRUE + plays_FALSE),
         ttl_dropbacks_wo_sacks = plays_FALSE + plays_TRUE) %>% 
  select(percent_play_action, everything()) %>% 
  filter(ttl_dropbacks_wo_sacks >= 60) %>%
  arrange(-epa_per_play_TRUE))
```

```{r}
play_action_2024 %>% 
  ggplot(.,
         aes(x=epa_per_play_FALSE,
             y=epa_per_play_TRUE,
             alpha=if_else(posteam == 'PIT', 'PIT', 'Other'))) +  
  geom_point(aes(size=plays_TRUE)) + 
  ggrepel::geom_text_repel(aes(label=passer),
                           family='Lexend', box.padding = .5) + 
  geom_hline(yintercept = 0, color='black', linetype='dashed', linewidth=.8) + 
  geom_vline(xintercept = 0, color='black', linetype='dashed', linewidth=.8) + 
  scale_alpha_manual(values = c('PIT' = 1,
                                'Other' = .2)) + 
  labs(x='EPA per play: NOT play action',
       y='EPA per play: IS play action') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
```

```{r top_10_perc_play_action_plot, include=TRUE}
play_action_2024 %>% 
  slice_max(percent_play_action, n=10) %>% 
  ggplot(., 
         aes(x = reorder(passer_id, -percent_play_action), 
             y = percent_play_action)) +
  geom_col(aes(color=posteam,
               fill=posteam)) +
  with_outer_glow(geom_text(aes(label = passer),
                            position = position_stack(vjust = .98),
                            angle = 90, hjust = .98, color = "white",
                            family = "Roboto",
                            fontface = "bold", size = 8),
                  sigma = 6, expand = 1, color = "black") +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::percent_format(),
                     expand = c(0,0)) +
  labs(x='',
       y='Play Action Rate (%)',
       title = "**Top 10 QBs in play action rate in 2024**",
       subtitle = "*Excludes QB Scrambles, minimum 60 dropbacks*",
       caption = "Source: {nflverse} package<br>and FTN charting data<br>Inspo: Brad Congelio") +
  nfl_analytics_theme() + 
  scale_fill_nfl(type = 'primary') + 
  scale_color_nfl(type = 'secondary') + 
  # adds headshots
  theme(axis.text.x = nflplotR::element_nfl_headshot(size = 1))
```

```{r top_10_epa_per_play_playaction_plot, include=TRUE}
play_action_2024 %>% 
  slice_max(epa_per_play_TRUE, n=10) %>% 
  ggplot(., 
         aes(x = reorder(passer_id, -epa_per_play_TRUE), 
             y = epa_per_play_TRUE)) +
  geom_col(aes(color=posteam,
               fill=posteam)) +
  with_outer_glow(geom_text(aes(label = passer),
                            position = position_stack(vjust = .98),
                            angle = 90, hjust = .98, color = "white",
                            family = "Roboto",
                            fontface = "bold", size = 7),
                  sigma = 6, expand = 1, color = "black") +
  geom_text(aes(label = plays_TRUE),
            position = position_stack(vjust = .05),
            color = "white", family = "Roboto", fontface = "bold", size = 3) + 
  annotate("text", label='Attempts', x=1, y=.06, 
           color='white', family = "Roboto", fontface = "bold", size = 3) + 
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     expand = c(0,0)) +
  labs(x='',
       y='EPA per play',
       title = "**Top 10 QBs in EPA per play on play action in 2024**",
       subtitle = "*Excludes QB scrambles, minimum 20 play action attempts<br>Through Week 8*",
       caption = "Source: {nflverse} package<br>and FTN charting data<br>Inspo: Brad Congelio") +
  nfl_analytics_theme() + 
  scale_fill_nfl(type = 'primary') + 
  scale_color_nfl(type = 'secondary') + 
  # adds headshots
  theme(axis.text.x = nflplotR::element_nfl_headshot(size = 1))
```




## Pat Mahomes' interception misfortune in 2024 -- what do our priors tell us about his likely performance from here? 

```{r passing_epa, include=TRUE}
(passing_epa <- pbp_plus_ftn_2022_24 %>% 
  filter(season == 2024) %>% 
  filter(!is.na(passer)) %>% 
  group_by(passer, posteam) %>% 
  summarise(plays = n(),
            count_drops = sum(is_drop, na.rm = TRUE),
            interceptions = sum(interception, na.rm = TRUE),
            sum_qb_epa = sum(qb_epa, na.rm = TRUE),
            sum_qb_epa_drops = sum(qb_epa[is_drop == TRUE], na.rm = TRUE),
            sum_qb_epa_int = sum(qb_epa[interception == 1], na.rm = TRUE),
            sum_qb_epa_sack = sum(qb_epa[sack == 1], na.rm = TRUE),
            sum_qb_epa_incomplete = sum(qb_epa[incomplete_pass == 1], na.rm = TRUE),
            sum_qb_epa_dpi = sum(qb_epa[penalty_type == 'Defensive Pass Interference'], na.rm = TRUE),
            .groups = 'drop') %>% 
  filter(plays >= 25) %>% 
  mutate(
    across(
      starts_with("sum"),
      list(
        # mean = ~round(mean(., na.rm = TRUE), 2),
        # zscore = ~scale(.),
        mean_centered = ~. - mean(., na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    )) %>% 
  arrange(-sum_qb_epa_int))
```

```{r}
passing_epa %>% 
  ggplot(.,
         aes(x=interceptions,
             y=sum_qb_epa_int,
             alpha=if_else(passer == 'P.Mahomes', 'P.Mahomes', 'Other'))) +
  geom_point() +
  ggrepel::geom_text_repel(aes(label = passer),
                           box.padding = .5) + 
  scale_x_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_y_continuous(breaks = scales::pretty_breaks(),
                     labels = scales::comma_format()) +
  scale_alpha_manual(values = c('P.Mahomes' = 1,
                                'Other' = .2)) + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
```

```{r player_stats_17_24, include=TRUE}
player_stats_17_24 <- load_player_stats(seasons = 2017:2024,
                                        stat_type = 'offense')
```


```{r}
(int_rate_17_24 <- player_stats_17_24 %>% 
  filter(position == 'QB') %>% 
  group_by(recent_team, player_name, season) %>% 
  summarise(
    across(
      c(attempts, interceptions),
      list(
        sum = ~sum(., na.rm = TRUE)
      ),
      .names = "{.fn}_{.col}"
    ),
    .groups = 'drop') %>% 
  mutate(int_rate = sum_interceptions / sum_attempts) %>% 
  filter(sum_attempts >= 100))
```


```{r}
int_rate_17_24 %>% 
  filter(player_name == 'P.Mahomes') %>% 
  ggplot(.,
         aes(x=season,
             y=int_rate)) + 
  geom_col() + 
  geom_mean_lines(aes(y0=int_rate)) + 
  geom_text(aes(label = paste(round(int_rate*100, 1), '%', sep='')),
            nudge_y = -.002, color='white', family='Lexend', fontface='bold') + 
  scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = seq(2017, 2024, 1)) +
  labs(x='',
       y='INT rate') + 
  ggtitle('Pat Mahomes has **doubled** his career average interception rate',
          subtitle = 'INT rate = interceptions / attempts<br>Does not exclude throw aways') + 
  nfl_analytics_theme()
```

```{r int_rate_jitter, include=TRUE, fig.width=10, fig.height=6.2}
int_rate_jitter <- int_rate_17_24 %>% 
  filter(season >= 2018) %>% 
  group_by(season) %>% 
  mutate(rank_int_rate = rank(int_rate),
         is_mahomes = if_else(player_name == 'P.Mahomes', 'P.Mahomes', 'Other')) %>% 
  ggplot(.,
         aes(x=season,
             y=int_rate,
             color=is_mahomes,
             alpha=is_mahomes)) + 
  # geom_jitter(height = 0) + 
  geom_jitter_interactive(aes(data_id = player_name, tooltip = player_name),
                          height = 0) + 
  ggtitle(glue("<span style='color:#E31837'>**Pat Mahomes**</span> usually boasts some of the lowest INT rates<br>in the NFL -- not in 2024")) + 
  scale_color_manual(values = c('P.Mahomes' = '#E31837',
                                'Other' = 'black')) + 
  scale_alpha_manual(values = c('P.Mahomes' = 1,
                                'Other' = .2)) + 
  scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = seq(2017, 2024, 1)) +
  labs(x='',
       y='INT rate') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')
```


```{r}
interactive_plot <- girafe(
  ggobj = int_rate_jitter
)
```

```{r int_rate_jitter_interactive, include=TRUE, fig.width=10, fig.height=6.2}
int_rate_jitter <- int_rate_17_24 %>% 
  filter(season >= 2018) %>% 
  group_by(season) %>% 
  mutate(rank_int_rate = rank(int_rate),
         is_mahomes = if_else(player_name == 'P.Mahomes', 'P.Mahomes', 'Other'),
         player_name_clean = stringr::str_replace_all(player_name, "'", "")) %>%  # Remove single quotes
  ggplot(aes(x = season,
             y = int_rate,
             color = is_mahomes,
             alpha = is_mahomes)) + 
  geom_jitter_interactive(aes(
    data_id = player_name_clean, 
    tooltip = glue("{player_name_clean} ({season}): {scales::percent(int_rate)}")
  ), height = 0) + 
  ggtitle(glue("Pat Mahomes usually boasts some of the lowest INT rates
in the NFL -- not in 2024")) + 
  scale_color_manual(values = c('P.Mahomes' = '#E31837', 'Other' = 'black')) + 
  scale_alpha_manual(values = c('P.Mahomes' = 1, 'Other' = 0.2)) + 
  scale_y_continuous(labels = scales::percent) + 
  scale_x_continuous(breaks = seq(2017, 2024, 1)) +
  labs(x = '', y = 'INT rate') + 
  nfl_analytics_theme() + 
  theme(legend.position = 'none')

interactive_plot <- girafe(ggobj = int_rate_jitter)

interactive_plot
```