---
title: "Analyzing Receiver Tracking Metrics from ESPN"
output: html_notebook
---

```{r libraries, include=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(rvest)
library(janitor)
library(httr)
library(jsonlite)
library(nflverse)
library(gt)
library(gtExtras)
library(ggtext)
library(ggrepel)
library(patchwork)
library(ggimage)
library(dlookr)
```

```{r nfl_analytics_theme, include=TRUE}
nfl_analytics_theme <- function(..., base_size = 12) {
  
  theme(
    text = element_text(family = "Lexend",
                        size = base_size,
                        color = "black"),
    axis.ticks = element_blank(),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16,
                                  vjust = .02,
                                  hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold", hjust = .5),
    strip.text = element_text(face = "bold"))
}
```

```{r teams, include=FALSE}
teams <- nflreadr::load_teams(current = TRUE)
```

```{r url, include=TRUE}
url <- 'https://espnanalytics.com/receivers'
```

```{r fetch_rtn_data, include=TRUE}
response <- GET("https://nfl-player-metrics.s3.amazonaws.com/rtm/rtm_data.json")
receiver_tracking <- fromJSON(rawToChar(response$content))
```

# Exploratory analysis of collected data 

```{r}
receiver_tracking %>% 
  dlookr::diagnose()
```

There are several columns that need further investigation, as it's unclear what the grain of the data may be: 

*  `min_season`, `max_season`, `min_player_season`, and `max_player_season` aren't clear. If you take one player (I'll use George Pickens below), it makes this point even muddier, as there are 10 records here for a player in his third season.

```{r}
receiver_tracking %>% 
  # this filtering should get 2024 data
  # filter(min_season == max_season & max_season == min_player_season & min_player_season == max_player_season & max_player_season == 2024) %>% 
  # this is including all seasons
  filter(min_season == max_season & max_season == min_player_season & min_player_season == max_player_season) %>% 
  mutate(yards_per_route_run = round(yds / rtm_routes, 1)) %>% 
  select(yards_per_route_run, max_player_season, full_nm, everything()) %>% 
  arrange(-yards_per_route_run)

```

```{r}
receiver_tracking %>% 
  filter(full_nm == 'George Pickens') %>% 
  filter(min_season == max_season & max_season == min_player_season & min_player_season == max_player_season) %>% 
  mutate(yards_per_route_run = round(yds / rtm_routes, 2),
         targets_per_route_run = round(rtm_targets / rtm_routes, 2)) %>% 
  select(yards_per_route_run, targets_per_route_run, everything())
```

```{r}
receiver_tracking %>% 
  filter(min_season == max_season & max_season == min_player_season & min_player_season == max_player_season & max_player_season == 2024) %>% 
  rename(overall_score = overall) %>% 
  pivot_longer(cols = ends_with('_score'),
               names_to = 'score') %>% 
  ggplot(.,
         aes(x=value,
             y=score,
             alpha=if_else(tm == 'PIT', 'PIT', 'Other'),
             color=if_else(tm == 'PIT', 'PIT', 'Other'))) + 
  geom_point() +
  scale_alpha_manual(values = c('PIT' = 1,
                                'Other' = .2)) + 
  scale_color_manual(values = c('PIT' = 'gold',
                                'Other' = 'black')) + 
  nfl_analytics_theme() +
  theme(legend.position = 'none')

?pivot_longer()
```



```{r player_stats_2024, include=TRUE}
player_stats_2024 <- load_player_stats(seasons = most_recent_season(),
                  stat_type = 'offense')
```

```{r analyze_player_stats_2024, include=TRUE}
player_stats_2024 %>% 
  filter(player_name == 'G.Pickens') %>% 
  group_by(player_name) %>% 
  rename(receiving_receptions = receptions,
         receiving_targets = targets) %>% 
  summarise(
    across(
      starts_with("receiving"),
      list(
        # mean = ~round(mean(., na.rm = TRUE), 2),
        sum = ~round(sum(., na.rm = TRUE), 2)
      ),
      .names = "{.fn}_{.col}"
    ))
```

I'll go to Pro Football Reference for the Average Depth of Target (ADOT) metric (and possibly others).

```{r player_ids, include=TRUE}
player_ids <- nflreadr::load_ff_playerids() %>% 
  filter(db_season == 2024) %>% 
  select(gsis_id, pfr_id) %>% 
  rename(player_id = gsis_id) %>% 
  filter(complete.cases(.))
# nflreadr::dictionary_ff_playerids
```

```{r pfr_stats_rec, include=TRUE}
pfr_stats_rec <- nflreadr::load_pfr_advstats(seasons = 2024, stat_type = "rec",
                                             summary_level = "season") %>% 
  left_join(x=.,
            y=player_ids,
            by='pfr_id')
```

```{r}
player_stats_2024 %>% 
  filter(player_name == 'G.Pickens') %>% 
  group_by(player_id, player_name) %>% 
  rename(receiving_receptions = receptions,
         receiving_targets = targets) %>% 
  summarise(
    across(
      starts_with("receiving"),
      list(
        sum = ~round(sum(., na.rm = TRUE), 2),
        mean = ~round(mean(., na.rm = TRUE), 3)
      ),
      .names = "{.fn}_{.col}"
    ),
    .groups = 'drop') %>% 
  left_join(x=.,
            y=pfr_stats_rec,
            by='player_id')
```


```{r}
pfr_stats_rec
```

## What are the Steelers missing in their WR room? 

WIP...

```{r}
receiver_tracking %>% 
  # this filtering should get 2024 data
  filter(min_season == max_season & max_season == min_player_season & min_player_season == max_player_season & max_player_season == 2024,
         tm == 'PIT')
```


