---
title: "Initial data analysis with nflverse packages"
output: html_notebook
---


```{r libraries, include=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nflverse)
```

```{r player_stats, include=TRUE}
player_stats <- nflreadr::load_player_stats(seasons = most_recent_season(),
                                            stat_type = "offense")
```

`nflreadr::load_player_stats()`. Arguments include:

* `seasons=`: numeric vector, or use the `most_recent_season()` function
* `stat_type=`: takes "offense", "defense", or "kicking"
*  `file_type=`: takes "rds" (default), "qs", "csv", or "parquet"


From the book:

> As an example, let’s say you need to get Patrick Mahomes’ total passing yard and attempts from the 2022 season. You could do so via load_pbp() but, if you do not need further context (such as down, distance, quarter, etc.), using load_player_stats() is much more efficient.

Below, I'm taking the weekly, box-score level receiving stats, and summarizing them for the pass catchers for the Steelers.

```{r}
player_stats %>% 
  filter(recent_team == 'PIT',
         position %in% c('WR', 'RB', 'TE')) %>% 
  rename(receiving_receptions = receptions,
         receiving_targets = targets) %>% 
  select(player_display_name, position, starts_with("receiving")) %>% 
  group_by(player_display_name, position) %>% 
  # summarize(across(starts_with("receiving"), mean, na.rm = TRUE))
  summarise(
    across(
      starts_with("receiving"),
      list(
        mean = ~round(mean(., na.rm = TRUE), 1),
        sum = ~round(sum(., na.rm = TRUE), 1)
      ),
      .names = "{.fn}_{.col}"
    ),
    .groups='drop'
  ) %>% 
  # ungroup() %>% 
  arrange(-sum_receiving_targets) %>% 
  filter(sum_receiving_targets >= 5)
  
```
  
```{r play_by_play_2024, include=TRUE}
play_by_play_2024 <- load_pbp(seasons=most_recent_season())
```
  

