---
title: "Initial data analysis with nflverse packages"
output: html_notebook
---


```{r libraries, include=TRUE, warning=FALSE, message=FALSE}
library(tidyverse)
library(nflverse)
library(gt)
```

```{r player_stats, include=TRUE}
player_stats <- nflreadr::load_player_stats(seasons = most_recent_season(),
                                            stat_type = "offense")
```

`nflreadr::load_player_stats()`. Arguments include:

* `seasons=`: numeric vector, or use the `most_recent_season()` function
* `stat_type=`: takes "offense", "defense", or "kicking"
*  `file_type=`: takes "rds" (default), "qs", "csv", or "parquet"


From the book:

> As an example, let’s say you need to get Patrick Mahomes’ total passing yard and attempts from the 2022 season. You could do so via load_pbp() but, if you do not need further context (such as down, distance, quarter, etc.), using load_player_stats() is much more efficient.

Below, I'm taking the weekly, box-score level receiving stats, and summarizing them for the pass catchers for the Steelers.

```{r}
player_stats %>% names()
```


```{r receiving_stats, include=TRUE}
(receiving_stats <- player_stats %>% 
  filter(position %in% c('WR', 'RB', 'TE')) %>% 
  rename(receiving_receptions = receptions,
         receiving_targets = targets,
         receiving_racr = racr,
         receiving_air_yards_share = air_yards_share,
         receiving_target_share = target_share,
         receiving_wopr = wopr) %>% 
  select(recent_team, player_display_name, position, starts_with("receiving")) %>% 
  group_by(recent_team, player_display_name, position) %>% 
  # summarize(across(starts_with("receiving"), mean, na.rm = TRUE))
  summarise(
    across(
      starts_with("receiving"),
      list(
        mean = ~round(mean(., na.rm = TRUE), 2),
        sum = ~round(sum(., na.rm = TRUE), 2)
      ),
      .names = "{.fn}_{.col}"
    ),
    games_played = n(),
    .groups='drop'
  ) %>% 
  # ungroup() %>% 
  arrange(-sum_receiving_targets) %>% 
  filter(sum_receiving_targets >= 5))
  
```

```{r}
receiving_stats %>% 
  group_by(recent_team) %>% 
  slice_max(sum_receiving_wopr) %>% 
  arrange(-sum_receiving_wopr)
```


`nflreadr::load_rosters()` is a way to pull in the headshots of players, which I'll use to make a nice `gt::` table of Pittsburgh's receiving in 2024.

```{r roster_2024, include=TRUE}
roster_2024 <- nflreadr::load_rosters(2024) %>%
  select(full_name, gsis_id, headshot_url)
```
  
 
  
Some of these statistics won't have any meaning summed (or possibly by taking an average), so I'll dig into each one to get clearer on that.
  
This looks like a good article on [EPA](https://bestballstats.com/expected-points-added-a-full-explanation/#:~:text=Expected%20Points%20Added%20(EPA)%20is,evaluate%20team%20and%20player%20performance.).

# Adjusted net yards per passing attempt 

```{r}
player_stats %>% 
  # add opponent_team to the group_by() for game-by-game stats
  group_by(recent_team, player_name) %>% 
  # this is copied from the book, which replicates the passing tables found in Pro Football Reference
  # the AY / A stat is the one that I'm most interested in
  summarize(
    completions = sum(completions),
    attempts = sum(attempts),
    cmp_pct = completions / attempts,
    yards = sum(passing_yards),
    touchdowns = sum(passing_tds),
    td_pct = touchdowns / attempts * 100,
    interceptions = sum(interceptions),
    int_pct = interceptions / attempts * 100,
    first_down = sum(passing_first_downs),
    yards_attempt = yards / attempts,
    adj_yards_attempt = (yards + 20 * touchdowns - 45 * interceptions) / attempts,
    yards_completions = yards / completions,
    yards_game = yards / 17,
    sacks = sum(sacks),
    sack_yards = sum(sack_yards),
    sack_pct = sacks / (attempts + sacks) * 100,
    .groups = 'drop') %>% 
  filter(attempts > 100) %>% 
  select(recent_team, player_name, adj_yards_attempt, everything()) %>% 
  # arrange(-adj_yards_attempt)
  arrange(-sack_yards)
```



## Play by play data

```{r play_by_play_2024, include=TRUE}
play_by_play_2024 <- load_pbp(seasons=most_recent_season())
```

## Average field position

```{r}
play_by_play_2024 %>% 
  # filter(posteam == 'PIT') %>%
  filter(!is.na(down)) %>% 
  group_by(game_id, drive) %>% 
  # Check that slicing takes the first play of each possession
  slice(1) %>% 
  # mutate(play_seq = n()) %>% 
  # select(play_seq, everything())
  ungroup() %>% 
  group_by(posteam) %>% 
  summarise(avg_starting_field_position = 100 - round(mean(yardline_100), 0)) %>% 
  mutate(rank_avg_starting_field_position = dense_rank(-avg_starting_field_position)) %>% 
  arrange(rank_avg_starting_field_position)
```


## Run-Pass split

Here's run-pass split, but it lacks in-game context. If a team is behind, they'll be much more likely to be pass-heavy.

```{r run_pass_split1, include=TRUE}
play_by_play_2024 %>% 
  filter(play_type %in% c('run', 'pass')) %>% 
  # could also add quarter
  group_by(posteam, play_type) %>% 
  tally() %>% 
  mutate(perc_total = round((n / sum(n)) * 100, 1)) %>% 
  arrange(-perc_total)
```

```{r run_pass_split2, include=TRUE}
play_by_play_2024 %>% 
  arrange(-score_differential) %>% 
  # probably an established way to get at this that takes into account time remaining in the game, too
  mutate(within_two_scores = if_else(abs(score_differential) <= 16, 1, 0)) %>% 
  filter(play_type %in% c('run', 'pass'),
         within_two_scores == 1) %>% 
  group_by(posteam, within_two_scores, play_type) %>% 
  tally() %>% 
  mutate(perc_total = round((n / sum(n)) * 100, 1)) %>% 
  arrange(posteam, -perc_total) %>% 
  gt() %>%
  cols_label(play_type = 'Play type',
             n = '# of plays',
             perc_total = '% of total') %>%
  #add table title
  tab_header(title = md("**Run-pass split by NFL teams in the 2024 Season**"),
             subtitle = 'Of plays run within 16 points (ahead or behind)') %>% 
  tab_source_note(source_note = "Data from `nflreadr::` package") %>% 
  #apply new style to all column headers
  tab_style(
    locations = cells_column_labels(columns = everything()),
    style = list(
      #thick border
      cell_borders(sides = "bottom", weight = px(3)),
      #make text bold
      cell_text(weight = "bold")
    )
  ) %>% 
  data_color(columns = c(perc_total),
             colors = viridis::magma(100)) %>% 
  #apply different style to title
  # tab_style(
  #   style = list(
  #     cell_fill(color = "lightblue"),   # Background color
  #     cell_text(weight = "bold")        # Bold text
  #   ),
  #   locations = cells_body(
  #     rows = posteam == "PIT"
  #   )
  # ) %>% 
  opt_all_caps() %>% 
  opt_table_font(
    font = list(
      google_font("Roboto"),
      default_fonts()
    )
  ) %>% 
  tab_options(
    #remove border between column headers and title
    column_labels.border.top.width = px(3),
    column_labels.border.top.color = "transparent",
    #remove border around the table
    table.border.top.color = "transparent",
    table.border.bottom.color = "transparent",
    #adjust font sizes and alignment
    source_notes.font.size = 12,
    heading.align = "left"
  )
```


## Average years per rush / pass play

## Personnel groupings 

## When is George Pickens targeted? 

### Splits

### Downs 

### Quarters 

### Position at snap 

### Redzone 


```{r}
play_by_play_2024 %>%
  filter(play_type == 'pass',
         posteam == 'PIT') %>% 
  # filter(receiver_player_name == 'Di.Johnson') %>%
  # select(touchdown, everything()) 
  # select(contains('catch'))
  mutate(redzone_play = if_else(yardline_100 <= 20, 1, 0),
         short_redzone_play = if_else(yardline_100 <= 10, 1, 0)) %>% 
  # select(touchdown, redzone_play, yardline_100, everything())
  filter(redzone_play == 1) %>% 
         # receiver_player_name == 'Di.Johnson',
         # need to figure out these fields. When I filter out NULL receiver player name, the receiving stats change. Look at Dionte Johnson (he has a receiving TD when I don't filter, but does not when I do)
  # when receiver_player_name IS NULL, it appears to be 1) a sack, or 2) an incompletion that is not a target (throwaway or bat down at the LOS)
         # !is.na(receiver_player_name)) %>%
  group_by(posteam, redzone_play, receiver_player_name) %>% 
    # need to add completions...
  summarise(count_targets = n(),
            sum_yards_gained = sum(yards_gained),
            sum_touchdown = sum(touchdown),
            sum_yards_after_catch = sum(yards_after_catch, na.rm = TRUE),
            .groups='drop') %>% 
  arrange(-count_targets)
  
group_by(pass_length, pass_location, game_half, no_huddle)
receiving_years
```


### Run success rate 

```{r run_success_rate, include=TRUE}
play_by_play_2024 %>% 
  filter(play_type == 'run',
         play == 1,
         # an NA rusher indicates a qb_scramble, so I'll filter those out of this
         qb_scramble != 1, 
         !is.na(down)) %>% 
  group_by(posteam, rusher) %>% 
  summarise(carries = n(), 
            success_rate = mean(success),
            .groups = 'drop') %>% 
  filter(carries >= 20) %>% 
  arrange(-success_rate) %>% 
  mutate(rank_success_rate = rank(-success_rate))
```

I'm curious about the success rate of Fields' runs, whether they were designed or a scramble. It doesn't look like the `qb_scramble` field is coded with any scrambles to date, so I'll let that one simmer for a few weeks.

```{r}
play_by_play_2024 %>% 
  # marking play_type = 'run' removes QB kneels 
  filter(play_type == 'run',
         rusher == 'J.Fields')
```


# From the book

```{r teams, include=TRUE}
teams <- nflreadr::load_teams(current = TRUE) %>%
  select(team_abbr, team_nick, team_name, team_color, team_color2)

teams
```

### Exercises

```{r pbp_2022, include=TRUE}
pbp_2022 <- load_pbp(2022) %>% 
  filter(season_type == 'REG')
```

1. Use filter() to select only Patrick Mahomes and all passes that resulted in interceptions.

```{r}
pbp_2022 %>% 
  filter(passer == 'P.Mahomes',
         interception == 1)
```


2. Add a summarize() to determine the average air yards (air_yards) for his combined interceptions.

```{r}
pbp_2022 %>% 
  filter(passer == 'P.Mahomes',
         interception == 1) %>% 
  group_by(posteam, passer) %>% 
  summarise(avg_air_yards = round(mean(air_yards, na.rm = TRUE), 1))
```

2.7.2 Exercise 2 

1. In a data frame titled wide_receivers, find the average yards after catch (yards_after_catch) and average yards after catch EPA (yac_epa) for the top-10 receivers in total receiving yards. In doing so, you will need to use group_by() on receiver.


```{r}
pbp_2022 %>% 
  filter(!is.na(receiver)) %>% 
  group_by(posteam, receiver) %>% 
  summarise(ttl_receiving_yards = sum(yards_gained, na.rm = TRUE),
            avg_yac = mean(yards_after_catch, na.rm = TRUE),
            avg_yac_epa = mean(yac_epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  top_n(10, ttl_receiving_yards)
```

2.7.3 Exercise 3 
1. In a data frame titled qb_hits, find the total number of times each defensive unit (defteam) recorded a QB hit (qb_hit).

```{r}
pbp_2022 %>% 
  filter(!is.na(defteam)) %>% 
  group_by(defteam) %>% 
  summarise(ttl_qb_hits = sum(qb_hit, na.rm = TRUE)) %>% 
  arrange(-ttl_qb_hits)
```

2.7.4 Exercise 4 
1. After running the following code, add a column titled difference using that mutate() verb that calculates the difference between each quarterback’s total_completions and total_incompletions.

```{r}
pbp_2022 %>%
  filter(complete_pass == 1 |
           incomplete_pass == 1 |
           interception == 1, !is.na(down)) %>%
  group_by(posteam, passer) %>%
  summarize(total_completions = sum(complete_pass == 1,
                                    na.rm = TRUE),
            total_incompletions = sum(incomplete_pass == 1,
                                      na.rm = TRUE),
            .groups = 'drop') %>%
  filter(total_completions >= 180) %>% 
  mutate(difference = total_completions - total_incompletions)
```

