---
title: "Cutting PBP for weekly analysis of effectiveness"
output: html_document
---

# Setup

```{r setup, include=TRUE, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(
    tidyverse,
    nflverse,
    gt,
    gtExtras,
    gtUtils,
    ggfx,
    rvest,
    httr,
    jsonlite,
    patchwork,
    ggtext,
    ggrepel,
    ggimage,
    dlookr,
    showtext,
    sysfonts,
    janitor,
    grid,
    ggiraph,
    glue,
    here
)    

sysfonts::font_add_google("Lexend")
showtext::showtext_auto()
```

```{r nfl_analytics_theme, include=TRUE}
nfl_analytics_theme <- function(..., base_size = 12) {
  theme(
    text = element_text(family = "Lexend", size = base_size, color = "black"),
    axis.ticks = element_blank(),
    axis.title = element_markdown(face = "bold"),
    axis.text = element_markdown(face = "bold"),
    plot.title.position = "plot",
    plot.title = element_markdown(size = 16, vjust = .02, hjust = 0.5),
    plot.subtitle = element_markdown(hjust = 0.5),
    plot.caption = element_markdown(size = 8),
    panel.grid.minor = element_blank(),
    panel.grid.major =  element_line(color = "#d0d0d0"),
    panel.background = element_rect(fill = "#f7f7f7"),
    plot.background = element_rect(fill = "#f7f7f7"),
    panel.border = element_blank(),
    legend.background = element_rect(color = "#F7F7F7"),
    legend.key = element_rect(color = "#F7F7F7"),
    legend.title = element_text(face = "bold", hjust = .5),
    strip.text = element_text(face = "bold"))
}
```

```{r teams, include=TRUE}
# teams <- nflreadr::load_teams(current = TRUE)
```

```{r pbp_2022_25, include=TRUE}
pbp_2022_25 <- load_pbp(seasons=2022:2025) %>% 
  filter(season_type == 'REG')
```

FTN's charting (which is available in the {nflverse} package), contains additional play-level information that can be added to the `load_pbp()` data.

```{r ftn_charting_2022_25, include=TRUE}
ftn_charting_2022_25 <- load_ftn_charting(seasons = 2022:2025)
```

```{r pbp_plus_ftn_2022_25, include=TRUE}
pbp_plus_ftn_2022_25 <- pbp_2022_25 %>% 
  left_join(x=.,
            y=ftn_charting_2022_25,
            by=c('game_id' = 'nflverse_game_id',
                 'play_id' = 'nflverse_play_id',
                 'season',
                 'week'))
```

## Weekly cuts (in progress)

```{r week_plot, include=TRUE}
week_plot <- pbp_plus_ftn_2022_25 %>% filter(season == 2025) %>% summarise(max(week)) %>% pull()
```

**Here, I'm thinking of setting up a structure that will allow for quick cuts of weekly data across many dimensions. Just starting to think through that structure.**

```{r}
pbp_plus_ftn_2022_25 %>% 
  # select(contains('is_'))
  filter(season == 2025,
         !is.na(passer),
         play_type == 'pass',
         is_no_huddle == TRUE) %>% 
  group_by(posteam, passer, is_no_huddle) %>% 
  summarise(mean_epa = mean(epa, na.rm = TRUE),
            plays = n(), 
            .groups = 'drop') %>% 
  filter(plays >= 10) %>% 
  mutate(mean_epa_rank = rank(-mean_epa)) %>% 
  arrange(mean_epa_rank)
```

```{r}
boolean_vars <- c("is_no_huddle", "is_motion", "is_play_action", 
                  "is_screen_pass", "is_rpo", "is_qb_out_of_pocket")
                  
double_vars <- c("down", "n_blitzers")

summary_vars <- c("play_type", "pass_length", "run_gap", "run_location")

# "is_trick_play", "is_interception_worthy", "is_throw_away", 
                  # "is_catchable_ball", "is_contested_ball", "is_created_reception", 
                  # "is_drop", "is_qb_sneak", "is_qb_fault_sack", 
```


```{r}
analyze_play_characteristics <- function(data, fields) {
  map_dfr(fields, function(field) {
    data %>%
      group_by(season, week, game_id, posteam, defteam, across(all_of(field))) %>%
      summarise(
        plays = n(),
        epa_per_play = mean(epa, na.rm = TRUE),
        success_rate = mean(success, na.rm = TRUE),
        .groups = "drop"
      ) %>%
      mutate(play_characteristic = field) %>%
      rename(characteristic_value = !!sym(field))
  })
}

test <- analyze_play_characteristics(pbp_plus_ftn_2022_25, summary_vars)

test %>% 
  filter(posteam == 'PIT',
         season == 2025,
         !is.na(characteristic_value),
         play_characteristic == 'is_play_action')


test %>% 
  filter(season == 2025, 
         week == 7,
         posteam == 'PIT',
         !is.na(characteristic_value)) %>% 
  arrange(play_characteristic, -epa_per_play)
```

Charting data appears to be NAs for no play situations, like ends of quarters.

```{r}
filter_posteam <- 'PIT'

pbp_plus_ftn_2022_25 %>% 
  filter(season == 2025,
         posteam == 'PIT') %>% 
  distinct(play_type)

plot_play_types <- function(data, pos_team = 'PIT', use_season = 2025, play_var){
  data %>% 
  filter(season == use_season, 
         posteam == pos_team,
         play_characteristic == play_var,
         !is.na(characteristic_value),
         # characteristic_value == TRUE,
         posteam == filter_posteam) %>% 
    ggplot(.,
         aes(x=epa_per_play,
             y=characteristic_value)) + 
    geom_boxplot() + 
    geom_jitter(aes(color=filter_posteam),
                size=1.5, width = 0, height = .05) + 
    labs(title=glue::glue(play_var, pos_team, use_season)) + 
    nfl_analytics_theme() + 
    scale_color_nfl(type = 'secondary')
}

plot_play_types(data = test, 
                use_season = 2025,
                pos_team = 'PIT',
                play_var = 'run_location')

```

```{r}
test %>% 
  filter(season == 2025, 
         # play_characteristic == 'is_play_action',
         !is.na(characteristic_value),
         # characteristic_value == TRUE,
         posteam == filter_posteam) %>% 
  ggplot(.,
         aes(x=epa_per_play,
             y=play_characteristic)) + 
  geom_boxplot() + 
  geom_jitter(aes(color=filter_posteam),
              size=1.5, width = 0, height = .05) + 
  nfl_analytics_theme() + 
  scale_color_nfl(type = 'secondary')
```


```{r}
pbp_plus_ftn_2022_25 %>% 
  # names()
  filter(game_id == '2025_18_BAL_PIT') %>% 
  # filter(is.na(is_play_action))
  group_by(posteam, is_play_action) %>% 
  summarise(plays = n(),
            epa_per_play = mean(epa, na.rm = TRUE),
            .groups = 'drop') %>% 
  filter(posteam == 'PIT')
```

This is not yet working as intended!

```{r}
pbp_plus_ftn_2022_25 %>% 
  filter(season == 2025,
         !is.na(passer)) %>% 
  group_by(posteam, passer) %>% 
  # summarise(across(boolean_vars))
  # mutate(
  summarise(
    n = n(),
    across(
      boolean_vars,
      list(
        mean = ~round(mean(epa, na.rm = TRUE), 2)
        # zscore = ~scale(.),
        # mean_centered = ~. - mean(., na.rm = TRUE)
      ),
      .names = c("mean_epa_{.col}"
    ))) %>% 
  filter(n >= 100) %>% 
  arrange(-mean_epa_is_no_huddle)
```